---
title: "Technical Appendix"
author: "Antonio Skilton"
output:
  html_notebook: 
    toc: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=8, fig.path='Figs/',
                      echo=T, warning=FALSE, message=FALSE)
```
##The Data

###Election Data
The data I am using is from [US Election Atlas](http://uselectionatlas.org/), which sells proprietary data. As a result, I will not include their entire dataset here. Instead, I will only include a variable that represents the vote difference between the candidates on a county level.
```{r, echo=T, message=F, warning=FALSE, eval=F}
library(tidyverse)
library(plotly)
```
```{r, echo=F, message=FALSE, warning=FALSE}
library(tidyverse, quietly=T)
library(plotly, quietly=T)
```

```{r, message=F, warning=F}
election <- read_csv("data/raw/Pres_Election_Data_2016i.csv", 
         na = "-", 
         col_types = cols_only(X1 = col_character(),
                               FIPS = col_character(),
                               ST = col_integer(),
                               X2 = col_character(),
                               Clinton = col_character(),
                               Trump = col_character(),
                               LSAD_TRANS = col_character())) %>% 
  filter(LSAD_TRANS %in% c("Parish", "County")) %>%
  slice(-1) %>% 
  rename(county_name = X1, 
         STATEABB = X2,
         COUNTY = FIPS,
         STATEFIPS = ST) %>% 
  mutate_at(3:4, funs(as.numeric(gsub("%", "", .))/100)) %>% 
  na.omit %>% 
  transmute(STATEFIPS, 
            county_name,
            STATEABB,
            COUNTY = as.numeric(substr(COUNTY, 3, 5)),
            vote_difference = Trump - Clinton,
            victor = ifelse(vote_difference > 0, "Trump", "Clinton")) 
```

"vote_difference" represents the margin of victory with which Donald Trump won a particular county. A positive value corresponds to a Trump victory, a negative value corresponds to a Clinton victory.
```{r, echo = F}
election
```

```{r, message=FALSE, warning=FALSE}
ggplot(election, aes(x = abs(vote_difference))) +
  geom_histogram() + 
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Histogram of Absolute Margins",
       subtitle = "The majority of counties were not competitive",
       x = "Absolute margin between Trump and Clinton") +
  theme_minimal()
```

```{r, message=FALSE, warning=FALSE}
ggplot(election, aes(x = vote_difference, fill = victor)) +
  geom_histogram() + 
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Histogram of Trump Margin of Victory",
       subtitle = "Most counties where Trump won were landslides. The opposite is true for Clinton.",
       x = "Trump margin of victory") +
  theme_minimal()
```


##Other county level data
There are county level datasets available for free from a variety of government sources. These were cleaned in R (HYPERLINK) then uploaded to SQL Server for further cleaning and data mining.

###County Density
from the [Census](http://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?src=bkmk)
```{r, message=F, warning=FALSE}
density <- read_csv("data/raw/county_density.csv", skip = 1, 
         col_types = cols_only(`Target Geo Id` = col_character(),
                               `Density per square mile of land area - Population` = col_double())) %>% 
  filter(nchar(`Target Geo Id`) == 14) %>% 
  transmute(STATEFIPS = as.numeric(substr(`Target Geo Id`, start = 10, stop = 11)),
            COUNTY = as.numeric(substr(`Target Geo Id`, start = 12, stop = 14)),
            people_per_sq_mile = `Density per square mile of land area - Population`)
```

```{r, echo=F, message=FALSE, warning=F}
density %>% 
  rename(`People per square mile` = people_per_sq_mile)
```


###Health insurance data 
from [Enroll America](https://www.enrollamerica.org/research-maps/maps/changes-in-uninsured-rates-by-county/):
```{r}
insurance <- read_csv("data/raw/County_Data_2016_health_insurance.csv",
         col_types = list(county_fips = col_character())) %>% 
  setNames(gsub(pattern = " ", replacement = "_", names(.))) %>% 
  transmute(STATEFIPS = ifelse(nchar(county_fips) == 5,
                               as.numeric(substr(county_fips, start = 1, stop = 2)),
                               as.numeric(substr(county_fips, start = 1, stop = 1))),
                               COUNTY = as.numeric(substr(county_fips, start = nchar(county_fips) - 2, stop = nchar(county_fips))),
                               uninsured_rate_2013 = `2013_uninsured_rate` / 100,
                               uninsured_rate_2016 = `2016_uninsured_rate` / 100,
                               decrease_from_2013_to_2016 = decrease_from_2013_to_2016 / 100) %>% 
  rename(change = decrease_from_2013_to_2016) %>% 
  arrange(STATEFIPS, COUNTY)
```

```{r, echo=F}
insurance %>% 
  rename(`Rate Uninsured in 2013` = uninsured_rate_2013,
         `Rate Uninsured in 2016` = uninsured_rate_2016)
```


###Economic data 
from the [USDA](https://www.ers.usda.gov/topics/rural-economy-population/rural-classifications.aspx).
```{r}
econ <- read_csv("data/raw/Unemployment.csv", skip = 6, 
         col_types = cols_only(FIPS_Code = col_character(),
                               Unemployment_rate_2015 = col_double(),
                               Median_Household_Income_2014 = col_number())) %>% 
  transmute(STATEFIPS = as.numeric(substr(FIPS_Code, 1, 2)),
            COUNTY = as.numeric(substr(FIPS_Code, 3, 5)),
            Unemployment_rate_2015 = Unemployment_rate_2015 / 100,
            Median_Household_Income_2014) %>% 
  filter(COUNTY != 0) 
```
```{r, echo=F}
econ %>% 
  rename(`Unemployment rate in 2015` = Unemployment_rate_2015,
         `Median household income in 2014` = Median_Household_Income_2014)
```


###County characteristics data 
from the [Census](https://www.census.gov/popest/data/counties/asrh/2015/CC-EST2015-ALLDATA.html):
```{r, message=FALSE, warning=FALSE}
NHWA <- read_csv("data/raw/CC-EST2015-ALLDATA.csv",
         col_types = cols_only(STATE = col_integer(),
                               COUNTY = col_integer(),
                               TOT_POP = col_integer(),
                               AGEGRP = col_integer(),
                               YEAR = col_integer(),
                               NHWA_MALE = col_integer(),
                               NHWA_FEMALE = col_integer()),
         progress = F) %>% 
  filter(AGEGRP >= 5,
         YEAR == 8) %>% 
  rename(STATEFIPS = STATE) %>%
  group_by(STATEFIPS, COUNTY) %>% 
  summarise(TOT_POP = sum(TOT_POP),
            NHWA = sum(NHWA_MALE + NHWA_FEMALE) / sum(TOT_POP))
```

```{r, echo=F}
NHWA %>% 
  rename(`Total Population (TOT_POP)` = TOT_POP,
         `Non-Hispanic White Alone (NHWA)` = NHWA)
```

###Education data 
from the [USDA](https://data.ers.usda.gov/reports.aspx?ID=18243):
```{r, message=FALSE, warning=FALSE}
education <- read_csv("data/raw/Education.csv", skip = 4) %>%
  slice(-c(1,2)) %>% 
  select(`FIPS Code`, contains("Percent")) %>% 
  setNames(gsub(pattern = "[[:punct:]]| ", 
                replacement = "_", names(.))) %>%
  setNames(gsub(pattern = "High_school_diploma", 
                replacement = "HSD", names(.), ignore.case = T)) %>%
  setNames(gsub(pattern = "Some_college_or_associate_s_degree|some_college_", 
                replacement = "someCollege", names(.), ignore.case = T)) %>%
  setNames(gsub(pattern = "bachelor_s_degree_or_higher", 
                replacement = "bachelorsOrMore", names(.), ignore.case = T)) %>%
  setNames(gsub(pattern = "with_less_than_a_HSD_|less_than_a_HSD_", 
                replacement = "lessThanHSD", names(.), ignore.case = T)) %>%
  setNames(gsub(pattern = "percent_of_", 
                replacement = "", names(.), ignore.case = T)) %>%
  setNames(gsub(pattern = "___|__", 
                replacement = "_", names(.))) %>%
  mutate(STATEFIPS = as.numeric(substr(FIPS_Code, 1, 2)),
         COUNTY = as.numeric(substr(FIPS_Code, 3, 5))) %>%
  select(STATEFIPS, COUNTY, -FIPS_Code, 
         adults_lessThanHSD_1970:adults_with_a_bachelorsOrMore_2010_2014) %>% 
  gather(key, value, -STATEFIPS, -COUNTY) %>% 
  transmute(STATEFIPS, COUNTY, key, percent = value / 100,
            year = gsub('\\D','\\1', key),
            year = as.numeric(substr(year, nchar(year) - 3, nchar(year))),
            year = ifelse(year == 2014, 2010, year)) %>% 
  spread(key, percent, drop = FALSE) %>% 
  unite(someCollege, contains("someCollege")) %>% 
  unite(bachelorsOrMore, contains("bachelorsOrMore"), contains("four_years_of_college_or_higher")) %>% 
  unite(lessThanHSD, contains("lessThanHSD")) %>% 
  unite(HSDonly, contains("HSD_only")) %>% 
  mutate_at(vars(bachelorsOrMore:HSDonly), funs(as.numeric(gsub("NA|_", "", .)))) %>% 
  filter(COUNTY != 0)
```

```{r, echo=F}
education
```

Using this time series data we can calculate growth rates. This is convenient because it returns us to a countly level dataset where each row represents a column, as opposed to a column-decade. Note that the growth rates are not annual growth rates but represent change over the course of 40 years.
```{r, message=FALSE, warning=FALSE}
education_growth <- read_csv("data/clean/Education.csv") %>% 
  group_by(STATEFIPS, COUNTY) %>% 
  summarise_at(vars(bachelorsOrMore:HSDonly), function(col) last(col) - first(col))
```

```{r, echo=F}
education_growth
```

```{r, fig.width=10, message=FALSE, warning=FALSE}
education_growth %>% 
  gather(variable, growth_rate, bachelorsOrMore:HSDonly) %>% 
  right_join(gather(education, variable, value, bachelorsOrMore:HSDonly)) %>% 
  unite(state_county, STATEFIPS, COUNTY) %>% 
  ggplot(aes(x = year, y = value, color = growth_rate, group = state_county)) +
    geom_line(alpha = 1/10) +
    theme(legend.position = "none") +
    facet_wrap(~variable) + 
    scale_y_continuous(labels = scales::percent) +
    scale_color_continuous(labels = scales::percent) +
    theme_minimal() + 
    labs(title = "Percent of Population by Year",
         y = "Percent of population")
```



Growth rates in each of the four variables in the education data have an independent effect on a vote for trump, as shown in the regression results below.
```{r, message=FALSE, warning=FALSE}
read_csv("data/clean/Education_growth.csv") %>% 
  lm(vote_difference ~ bachelorsOrMore + someCollege + lessThanHSD + HSDonly, data = .) %>% 
  summary
```


#Joined Data
I join all datasets:
```{r, message=FALSE, warning=FALSE}
joined_data <- election %>% 
  inner_join(education_growth) %>% 
  inner_join(density) %>% 
  inner_join(insurance) %>% 
  inner_join(econ) %>% 
  inner_join(NHWA)
```

```{r, echo=F, message=FALSE, warning=FALSE}
joined_data
```

###Histograms
```{r, fig.width=10}
joined_data %>%
  gather(key, value, bachelorsOrMore:NHWA) %>% 
  ggplot(aes(x = value)) +
    geom_histogram() +
    facet_wrap(~key, scales = "free") + 
    theme_minimal()
```

###Random Forest
The Random Forest model produces a list of variables in the order of importance in determining voter_difference:
```{r, message=FALSE, warning=FALSE}
library(randomForest)
data.rf <- randomForest(x = select(joined_data, -vote_difference, -victor),
                        y = joined_data$vote_difference,
                        importance = T)
imp <- importance(data.rf, 1)[order(importance(data.rf, 1), decreasing = T),] 
tibble(Variable = names(imp),`%Increase in MSE` = round(imp, 2) / 100)
```

###Correlations
The "vote_difference" variable is of greatest interest here. The following correlations show us which variables are most related to the election outcomes per county.
```{r, fig.height=5}
library(antonioSkiltonTools)
library(forcats)

joined_data %>% 
  rankedCorrelations() %>% 
  filter(Vars1 == "vote_difference" | Vars2 == "vote_difference") %>% 
  unite(Vars, Vars1, Vars2, sep = " & ") %>% 
  arrange(abs(correlation)) %>% 
  mutate(Vars = fct_inorder(Vars)) %>% 
  ggplot(aes(x = Vars, y = correlation, alpha = log(abs(correlation)))) +
    geom_bar(stat = "identity") + 
    coord_flip() + 
    theme_minimal() +
    scale_y_continuous(labels = scales::percent, limits = c(-1,1)) +
    scale_alpha(guide = F) +
    labs(title = "Ranked Correlations",
         y = "Degree of correlation",
         x = "Variable pairs")
```

To detect multicolinearity, we show correlations between independent variables.
```{r, fig.height=10}
joined_data %>% 
  rankedCorrelations() %>% 
  filter(Vars1 != "vote_difference" & Vars2 != "vote_difference") %>% 
  unite(Vars, Vars1, Vars2, sep = " & ") %>% 
  arrange(abs(correlation)) %>% 
  mutate(Vars = fct_inorder(Vars)) %>%
  slice(54:107) %>% 
  ggplot(aes(x = Vars, y = correlation)) +
    geom_bar(stat = "identity") + 
    coord_flip() + 
    theme_minimal() +
    scale_y_continuous(labels = scales::percent, limits = c(-1,1)) +
    labs(title = "Ranked Correlations",
         y = "Degree of correlation",
         x = "Variable pairs")
```
```{r, fig.height=10, echo=F}
joined_data %>% 
  rankedCorrelations() %>% 
  filter(Vars1 != "vote_difference" & Vars2 != "vote_difference") %>% 
  unite(Vars, Vars1, Vars2, sep = " & ") %>% 
  arrange(abs(correlation)) %>% 
  mutate(Vars = fct_inorder(Vars)) %>%
  slice(1:53) %>% 
  ggplot(aes(x = Vars, y = correlation)) +
    geom_bar(stat = "identity") + 
    coord_flip() + 
    theme_minimal() +
    scale_y_continuous(labels = scales::percent, limits = c(-1,1)) +
    labs(title = "Ranked Correlations",
         y = "Degree of correlation",
         x = "Variable pairs")
```
#Charts
```{r, fig.width=10}
p <- joined_data %>% 
  unite(Location, county_name, STATEABB, sep = ", ") %>% 
  ggplot(aes(x = NHWA, y = vote_difference, size = TOT_POP, color = victor, group = Location)) + 
    geom_point(alpha = 1/3) + 
    theme_minimal() +
    theme(legend.title = element_blank()) +
    scale_color_brewer(type = "qual", palette = "Set1", direction = -1) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) + 
    scale_size(guide = F) +
    labs(title = "Counties by Trump Victory Margin and Whiteness",
         x = "Percent of county that is non-Hispanic White",
         y = "Margin of Trump lead")

ggplotly(p, tooltip = c("group"))
```

####County Populations
```{r, fig.width=8, message=FALSE, warning=FALSE, fig.height=3}
joined_data %>% 
  ggplot(aes(x = TOT_POP / 1000, fill = victor)) + 
    geom_histogram(bins = 75) + 
    facet_wrap(~victor) +
    theme_minimal() +
    xlim(0,1000) +
    scale_y_log10() +
    scale_fill_brewer(type = "qual", palette = "Set1", direction = -1,
                      guide = F) +
    labs(title = "Distribution of County Populations",
         subtitle = "Counties in which Trump tend to be smaller.",
         x = "Total population in thousands")
```

####Whiteness
```{r, fig.width=8}
p2 <- joined_data %>% 
  unite(Location, county_name, STATEABB, sep = ", ") %>% 
  ggplot(aes(x = NHWA, y = vote_difference, size = TOT_POP, color = victor, group = Location)) + 
    geom_point(alpha = 1/3) + 
    facet_wrap(~cut(people_per_sq_mile, 
                    breaks = c(-Inf, 15, 300, Inf), 
                    labels= c("low density", "medium density", "high density")), scales = "free_x") +
    theme_minimal() +
    theme(legend.title = element_blank()) +
    scale_color_brewer(type = "qual", palette = "Set1", direction = -1) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) + 
    scale_size(guide = F) +
    labs(title = "Counties by Trump Victory Margin and Whiteness",
         subtitle = "Low density counties are more White and more pro Trump",
         x = "Percent of county that is non-Hispanic White",
         y = "Margin of Trump lead")

ggplotly(p2, tooltip = "group")
```

####College Education
```{r, fig.width=8}
p3 <- joined_data %>% 
  unite(Location, county_name, STATEABB, sep = ", ") %>% 
  ggplot(aes(x = bachelorsOrMore, y = vote_difference, size = TOT_POP, color = victor, group = Location)) + 
    geom_point(alpha = 1/3) + 
    theme_minimal() +
    theme(legend.title = element_blank()) +
    scale_color_brewer(type = "qual", palette = "Set1", direction = -1) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) + 
    scale_size(guide = F) +
    labs(title = "Trump Victory and College Education",
         subtitle = "Trump more likely to win counties with a lower share of college educated people.",
         x = "Percent of county that has at least a bachelors degree",
         y = "Margin of Trump lead")

ggplotly(p3, tooltip = "group")
```

####Health Insurance and Median Household Income
```{r, fig.width=10}
p4 <- joined_data %>% 
  unite(Location, county_name, STATEABB, sep = ", ") %>% 
  ggplot(aes(x = Median_Household_Income_2014, y = uninsured_rate_2013, size = TOT_POP, group = Location)) + 
    geom_point(alpha = 1/5) + 
    theme_minimal() +
    scale_x_continuous(labels = scales::dollar) + 
    scale_y_continuous(labels = scales::percent) +
    scale_size(guide = F) +
    labs(title = "Rate of Uninsured by Median Household Income",
         y = "Percent of county that is uninsured",
         x = "Median Household Income in 2014")

ggplotly(p4, tooltip = "group")
```


